
# Building a Data Package


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
```

In data science, it’s vital to have some useful data available. But it’s much better to have this data ready in a reproducible way and provide it in a simple data R package, so that your team can start using it right away.

Let’s make an example R data package in seven simple steps!

##  Install ‘devtools’ package in R
First, we need to install the {devtools} package. Start a new R session (or RStudio) and run:

```{r}
install.packages("devtools")
```

It comes along with package {usethis} and some handy and helper functions that will ease the creation of a new R package for you.

## Create a new package
Now it’s time to create a backbone of our new R package. This step is pretty straightforward, just run:

```{r}
usethis::create_package("v0te")
```


My package will be named “v0te” and will contain United Nations General Assembly Voting Data from the Harvard Dataverse.1

This command will create a new folder in your home directory prepopulated with some files for further adjustment.

## Edit DESCRIPTION file
The DESCRIPTION file contains some essential metadata about your package. It is already prefilled with few lines, so open it and fill in your name and a package description. For example:

```{r, eval=FALSE}
Package: v0te
Title: United Nations General Assembly Voting Data
Version: 0.0.0.9000
Authors@R:
    person(given = "Petr",
           family = "Kajzar",
           role = c("aut", "cre"),
           email = "author@example.org")
Description: The package contains UN General Assembly Voting Data. It is a
    dataset of roll-call votes in the UN General Assembly from 1946.
License: MIT + file LICENSE
Encoding: UTF-8
LazyData: true
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.1.1
Depends:
    R (>= 2.10)
    
```    


## Add your data
OK, we have metadata prepared, now we’ll add some real data. We now need to create a data directory to store the data for this package. First we will create a directory to store our scripts that will collect and process our raw data. In your R concole, type:

```{r, eval=FALSE}
usethis::use_data_raw("v0te")
```


This will create a data-raw folder with v0te.R file. Open that (if it’s not already open) and edit.

In my case, I will load remote data, edit that and save to data folder easily with usethis::use_data() function.

My data-raw/v0te.R file looks like this:

## load data
```{r, eval=FALSE}
load(url("https://dataverse.harvard.edu/api/access/datafile/:persistentId?persistentId=doi:10.7910/DVN/LEJUQZ/MLNAHB"))
```



## extract only selected columns
```{r, eval=FALSE}
votes <- subset(completeVotes, select = c("resid", "year", "Countryname", "vote"))
```


## rename columns
```{r, eval=FALSE}
colnames(votes) <- c("resid", "year", "country", "vote")
```


## change votes to factor and self-descriptive
```{r, eval=FALSE}
votes$vote <- as.factor(votes$vote)
levels(votes$vote) <- c("yes", "abstain", "no", "absent", "not member")
```


## use that data!
```{r, eval=FALSE}
usethis::use_data(votes, overwrite = TRUE)
```

After editing the file, it is necessary to run this code immediately! Run it either by selecting it in R GUI and pressing Ctrl+R, or in RStudio by clicking on “Run” button. When you run your code, R will process the data and save a new dataset to the data folder.

## Document your dataset
We have our data-raw folder ready - it reproducibly creates the dataset. The dataset is available in the data folder. So, now we only need to document that!

Run:

```{r}
usethis::use_package_doc()
```

in your R console and R will open a documentation file in R folder for your package.

You can use roxygen2 block for your documentation, see https://r-pkgs.org/data.html#documenting-data.

My R/v0te-package.R file looks like this:

```{r, eval=FALSE}

#' @keywords internal
"_PACKAGE"

# The following block is used by usethis to automatically manage
# roxygen namespace tags. Modify with care!
## usethis namespace: start
## usethis namespace: end
NULL

#' United Nations General Assembly Voting Data
#'
#' This is a dataset of roll-call votes in the UN General Assembly from 1946.
#'
#' @format A data frame with 4 variables:
#' \describe{
#'   \item{resid}{roll-call vote id}
#'   \item{year}{year of vote}
#'   \item{country}{country name}
#'   \item{vote}{yes/no/abstain/absent/not a member}
#' }
#' @source \url{https://doi.org/10.7910/DVN/LEJUQZ}
"votes"
```


The first two blocks were generated automatically, and I don’t touch that. The third block is written by me and contains basic info about the dataset.

## Generate documentation files
If you have your documentation ready, run:
```{r, eval=FALSE}
devtools::document()
```


This will generate documentation files for your package inside the man folder.

## Congratulations!
Your package is probably ready to use. Please, consider to add some permissive license, e.g. MIT:

```{r, eval=FALSE}
usethis::use_mit_license("Your Name")
```

And now you can test your package with:

```{r, eval=FALSE}
devtools::load_all()
```

You’ll see that you can use a new votes dataset and do some data science on that. :-)

```{r, eval=FALSE}
str(votes)
```

Lastly, check your package:

```{r, eval=FALSE}
devtools::check()
```

If everything looks good, just upload your package to some git hosting site.
